---
title: "convergence_analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{convergence_analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(papss)
```

## Simulate pupil data

Here we again start by simulating some pupil data. We only look at 10 subjects here
since we want to collect the history of coefficient updates this time, which quickly
results in very large matrices.

```{r}
n <- 10
sim_obj <- additive_pupil_sim(n_sub = n,
                              slope_dev = 1.5,
                              sub_dev = 0.15,
                              trial_dev = 0.25,
                              residual_dev=15.0,
                              should_plot=T)
dat <- sim_obj$data
```

## Aggregate
We again aggregate the raw data, as is explained further in the artificial_data_analysis vignette.

```{r}
aggr_dat <- aggregate(list("pupil"=dat$pupil),by=list("subject"=dat$subject,"time"=dat$time),FUN=mean)

aggr_dat <- aggr_dat[order(aggr_dat$subject),]

plot(aggr_dat$time[aggr_dat$subject == 1],
     aggr_dat$pupil[aggr_dat$subject == 1],
     type="l",ylim=c(min(aggr_dat$pupil),
                     max(aggr_dat$pupil)),
     xlab="Time",
     ylab="Pupil dilation",
     main= "Average pupil trajectories for all subjects",
     lwd=3)

for (si in 2:n) {
 sub_aggr_dat <- aggr_dat[aggr_dat$subject == si,]
 lines(sub_aggr_dat$time,sub_aggr_dat$pupil,col=si,lwd=3)
}

```

We also need to define pulse locations again.

```{r}
# Calculate pulse locations
last_pulse_dat <- length(unique(aggr_dat$time)) - (3 * round(930/100)) - 1
  
# Define possible pulse location
pulse_locations_dat <- seq(1,last_pulse_dat,1)
real_locations_dat <- unique(aggr_dat$time)[pulse_locations_dat]
```

## Investigating the identifiability of the problems
How will the solver deal with different starting values?

```{r}
iter <- 10
hist_collection <- list()

for(i in 1:iter){
  # Pick a new seed
  set.seed(i)
  
  # Solve the pupil spline
  solvedPupil <- papss::pupil_solve(pulse_locations_dat,
                             real_locations_dat,
                             data = aggr_dat,
                             maxiter_inner = 10000,
                             maxiter_outer = 10,
                             model="WIER_SHARED",
                             convergence_tol = 1e-06,
                             should_collect_progress=T)
  
  # Coefficient history
  hist = solvedPupil$coefHistory
  hist_collection[[i]] <- hist
}
```

## Coefficient changes over time

This looks very different - apart from the clearly identifiable slopes.

```{r}
for(hi in 1:iter){
  chist <- hist_collection[[hi]]
  plot(1:ncol(chist),
       chist[1,],
       type="l",
       lwd=3,
       ylim=c(-3,3),
       xlab="Iterations",
       ylab="Coefficient value")
  
  for(ri in 2:nrow(chist)){
    lines(1:ncol(chist), chist[ri,],lwd=3,col=ri)
  }
}
```


But what if we look only at the last column, i.e., the final estimates? Close agreement! 
```{r}
chist <- hist_collection[[1]]
plot(1:nrow(chist),
       chist[,ncol(chist)],
       ylim=c(-3,3),
       xlab="Coefficient",
       ylab="Coefficient value",type="l",lwd=3)

for(hi in 2:iter){
  chist <- hist_collection[[hi]]
  plot(1:nrow(chist), chist[,ncol(chist)],
       col=ri,
       ylim=c(-3,3),
       xlab="Coefficient",
       ylab="Coefficient value",type="l",lwd=3)


}
```